---
title: "Shifting Borders: Analyzing U.S. Immigration Trends Through the Years"
output: html_document
---



#INTRODUCTION: 

When you think about what makes America, **AMERICA**, certain thoughts come to mind. Most importantly our diverse population. Well, we didn’t all get here by magic. One person or a group of people, a family perhaps, migrated here for a multitude of reasons i.e. fleeing violence, poverty, and ultimately seeking opportunity. We are a nation of immigrants, and have been since our conception. 

Our project centers around this defining feature of American life by exploring immigration trends in the United States, using data collected by U.S. Customs and Border Protection's Office of Field Operations (OFO). Our focus begins in 2004, a pivotal year in the modern immigration landscape.

##Why does our data start in 2004? 
To grasp, the importance of our data we must paint the picture of what was going on in the United States immigration landscape, pre-2004. 
###**PRE-2004 IMMIGRATION LANDSCAPE:**

REVAMP THIS KNOWLEDGE!!
Mostly European immigrants… The Story of Immigration to the United States | FAIRstory #11…


However, security concerns following 9/11 in 2001 dramatically reshaped how the U.S. approached immigration enforcement. The creation of the Department of Homeland Security (DHS) in 2002 brought multiple immigration functions under one roof, with a sharp increase in border surveillance, administrative enforcement, and data collection.

###**2004, THE TURNING POINT:**

By 2004, new post-9/11 policies were fully in effect, reshaping how immigration was tracked, processed, and reported. That’s why most modern datasets — including the one we’re analyzing — begin here. It marks the start of a new era: one that emphasizes data-driven border enforcement, detailed classifications, and nuanced tracking by citizenship, family status, enforcement outcomes, and more.

###**POST–2004 IMMIGRATION LANDSCAPE:** 

From 2004 onward, immigration patterns fluctuated in response to political leadership, global conflicts, economic shifts, and humanitarian crises. For instance:

•	In 2015, the U.S. saw a surge in unaccompanied minors and family units from Central America.
•	The COVID-19 pandemic (2020–2021) disrupted migration drastically.
•	In recent years, increased migration from Venezuela, Haiti, and Nicaragua has reshaped the encounter landscape.
2015, Largest #Finish thought from video. 

##ABOUT OUR DATASET:

The dataset we’re analyzing is sourced from the KHSM Encounters (OFO) reports and contains encounter-level data tracked by:

•	Year/Month
•	Citizenship
•	Region
•	Family Status
•	Enforcement vs Administrative Actions
•	Field Office Location

##**Research questions:**

Why does our data start in 2004? 
What happened in 2004? 
#Questions that Chat helped with. 


1.	How have total border encounters changed over time (2004–present)?
2.	Which nationalities are most commonly encountered and how have they shifted over time?
3.	How does family status (e.g., single adult vs family unit) impact enforcement outcomes?
4.	Are there regional patterns in encounter trends — and which field offices process the highest volumes?

Questions that came up as we deepened our understanding of the dataset: adding to see how it affects my commit and push process.

___
---
Spaghetti Graph of 2004 - 2014.
```{r}
#Import libraries.

#Load ReadXL to be able to read Excel sheets.
library(readxl) 

#Load dplyr for data manipulation.
library(dplyr) 

#Load Ggplot to display data visualizations.
library(ggplot2) 

```


```{r}

#Define path to Excel file and load annual region data.
annual_region_encounters <- read_excel("KHSM Encounters (OFO) fy25m11.xlsx", sheet = "Annual Region")

print("A glimpsee into the data: ")
glimpse(annual_region_encounters)
print("___________________________")

print("Let's see the column names: ")
colnames(annual_region_encounters)

print("How many regions are there?")
length(unique(annual_region_encounters$Region))


#Pipe the dataframe into the distinct function so that it can pull all the unique regions. 
annual_region_encounters %>%
  distinct(Region)

```
Now we have an idea of what the dataset contains we want to visualize it. The plot of choice will be a spagetti plot because....

```{r}
# Clean and filter for 2004–2025 so that we can get a full picture of all the data. 
#Pipe in the data and rename the columns for future readability. 
spaghetti_annual_region_encounters_pre <- annual_region_encounters %>%
  rename(
    Year = `Fiscal\r\nYear`,
    Region = `Region`, 
    Encounters = `Quantity`
  ) %>%
  filter(Year >= 2004 & Year <= 2025)

length(unique(annual_region_encounters$Region))


top_regions <- spaghetti_annual_region_encounters_pre %>%
  group_by(Region) %>%
  summarise(Total = sum(Encounters)) %>%
  arrange(desc(Total)) 

spaghetti_annual_region_encounters_filtered_pre <- spaghetti_annual_region_encounters_pre %>%
  filter(Region %in% top_regions$Region)


print(spaghetti_annual_region_encounters_filtered_pre)
ggplot(spaghetti_annual_region_encounters_pre, aes(x = Year, y = Encounters, color = Region)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Spaghetti Plot: Border Encounters by Region (2004–2025)",
    x = "Year",
    y = "Encounters"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14)
  )
```
What this data is telling us? 

From what we know about the political landscape at the time, 9/11/2001 was the initial dominoe effect that....


Despite all that took place around 2015 we notice a spike in 2015, which is indicative of...

We also notice a sharp decline in 2020-2021, which so happens to coincide with the World Health Organization declaration of the COVID-19 pandemic. 

that decline can be seen in all borders except one, the coastal one. We took to the internet to answer why, while everyone, everyplace in the world seemed to stop, that boarders encounters not only stayed steady but seemed to increase...
From the data there seems to be a spike in the 

Let's take a closer look at 2014-2025. 


```{r}
# Clean and filter for 2004–2025 so that we can get a full picture of all the data. 
#Pipe in the data and rename the columns for future readability. 
spaghetti_annual_region_encounters_pre <- annual_region_encounters %>%
  rename(
    Year = `Fiscal\r\nYear`,
    Region = `Region`, 
    Encounters = `Quantity`
  ) %>%
  filter(Year >= 2014 & Year <= 2025)

length(unique(annual_region_encounters$Region))


top_regions <- spaghetti_annual_region_encounters_pre %>%
  group_by(Region) %>%
  summarise(Total = sum(Encounters)) %>%
  arrange(desc(Total)) %>%
  slice_head(n = 5)

spaghetti_annual_region_encounters_filtered_pre <- spaghetti_annual_region_encounters_pre %>%
  filter(Region %in% top_regions$Region)


print(spaghetti_annual_region_encounters_filtered_pre)
ggplot(spaghetti_annual_region_encounters_pre, aes(x = Year, y = Encounters, color = Region)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Spaghetti Plot: Border Encounters by Region (2014–2025)",
    x = "Year",
    y = "Encounters"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14)
  )
```

The next question to answer is: what is driving these stark fluctuations? 

The graphs reflects the policy changes that have taken place administration to administration. 

The most important changes that took place in the context of immigration policy were: 

•	2002: Creation of DHS
•	2004: Post-9/11 enforcement begins
•	2012: DACA introduced
•	2017: Trump’s executive orders
•	2020: Title 42 and COVID-19 emergency rules
•	2023: Changes under Biden (e.g., expanded TPS)


In order to clearly see the affects these policies had on the flunctuations and declines we have created a geom...graph that would  they have on the outco


```{r}
ggplot(filtered_full_data, aes(x = Year, y = Encounters, color = Region)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  
 
  geom_vline(xintercept = 2004, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2012, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2015, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2023, linetype = "dashed", color = "gray30") +

  annotate("text", x = 2004, y = max(filtered_full_data$Encounters)*0.95, 
           label = "DHS Begins", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2012, y = max(filtered_full_data$Encounters)*0.85, 
           label = "DACA", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2015, y = max(filtered_full_data$Encounters)*0.75, 
           label = "Family Surge", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2020, y = max(filtered_full_data$Encounters)*0.65, 
           label = "COVID + Title 42", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2023, y = max(filtered_full_data$Encounters)*0.55, 
           label = "Title 42 Ends", angle = 90, vjust = -0.4, size = 3) +

  labs(
    title = "Border Encounters by Region with Key Immigration Policy Events 
                                     (2004–2025)",
    x = "Year",
    y = "Number of Encounters"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14)
  )

```
Summary: 

Now that we put all these factors together it is 

Looking at something that seems so... to American 