---
title: "Shifting Borders: Analyzing U.S. Immigration Trends Through the Years"
output: html_document
---



# INTRODUCTION

When you think about what makes America, **AMERICA**, certain thoughts come to mind. Most importantly our diverse population. Well, we didn’t all get here by magic. One person or a group of people, a family perhaps, migrated here for a multitude of reasons i.e. fleeing violence, poverty, and ultimately seeking opportunity. We are a nation of immigrants, and have been since our conception. 

Our project centers around this defining feature of American life by exploring immigration trends in the United States, using data collected by U.S. Customs and Border Protection's Office of Field Operations (OFO). Our focus begins in 2004, a pivotal year in the modern immigration landscape.

## Why does our data start in 2004? 

To grasp, the importance of our data we must paint the picture of what was going on in the United States immigration landscape, pre-2004. 

### **PRE-2004 IMMIGRATION LANDSCAPE**

Long story short, the immigration story of America begins with the first settlers from Europe that were mostly fleeing to escape religious persecution. As decades and centuries past the "New Americans" started to find  way to keep certain people out while allowing others to make America home. Thus forming the infancy of what we know as our Immigration Policy. 

One example being the Chinese Exclusion Act of 1882 was past that excluded Chinese laborers from being able to migrate here for fear of that they would "take our jobs". Sounds familiar, doesn't it?

While purposely excluding others, the immigration policy of the United States at the time did favor other rational groups, mostly white, Anglo-Saxon, Protestant (W.A.S.P.) Europeans. 

Fast forward to now, maybe the sentiment of wanting to keep people out is just as American as they come. But over time, mostly after the Immigration and Nationality Act of 1965, the U.S. worked to remove it's racist racially motivated quotas. Official borders were created and America seemed more open to Asylum seekers. 

Things began to change on 9/11/2001. 

Immigration policy became synonymous with National Security. 

The Bush Administration created the Department of Homeland Security in 2002. And border control and enforcement became more aggressive, and militant. 


### **2004-THE TURNING POINT**
The reason our graph starts in 2004 is because it marks the beginning of a new era in U.S. immigration policy — one defined by heightened national security and data collection following the 9/11 attacks.

By 2004, new post-9/11 policies were fully in effect, reshaping how immigration was tracked, processed, and reported. 

### **POST–2004 IMMIGRATION LANDSCAPE** 

From 2004 onward, immigration patterns fluctuated in response to political leadership, global conflicts, economic shifts, and humanitarian crises. For instance:

-	In 2015, the U.S. saw a surge in unaccompanied minors and family units from Central America.
-	The COVID-19 pandemic (2020–2021) disrupted migration drastically.
-	In recent years, increased migration from Venezuela, Haiti, and Nicaragua has reshaped the encounter landscape.


## ABOUT OUR DATASET

The dataset we’re analyzing is sourced from the KHSM Encounters (OFO) reports and contains encounter-level data tracked by:

-	Fiscal Year
-	Region
- Quantity| Number of Encounters

## **Research questions**

How have total border encounters changed over time (2004–present)?

## Questions that came up while we were analyzing. 

- Why does our data start in 2004? 
- What happened in 2004?
- Which nationalities are most commonly encountered and how have they shifted over time?
- Are there regional patterns in encounter trends — and which field offices process the highest volumes?
- Can you paint the picture of these boarders and where they are located so that we can have more information on them? 
___
First things first, let's see the data!

After speaking with Dr. Shoemaker we decided to use a spagetti graph to display our data because we learned that is the best way to display and analyse our data. A spagetti graph would allow us to  visualize how border encounters have changed over time across multiple regions simultaneously.


Spaghetti Graph of 2004 - 2014.
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,        # Hides all code unless you override
  message = FALSE,     # Hides messages
  warning = FALSE      # Hides warnings
)
#Import libraries.

#Load ReadXL to be able to read Excel sheets.
library(readxl) 

#Load dplyr for data manipulation.
library(dplyr) 

#Load Ggplot to display data visualizations.
library(ggplot2) 

```


```{r}

#Define path to Excel file and load annual region data.
annual_region_encounters <- read_excel("KHSM Encounters (OFO) fy25m11.xlsx", sheet = "Annual Region")

print("A glimpsee into the data: ")
glimpse(annual_region_encounters)
print("___________________________")

print("Let's see the column names: ")
colnames(annual_region_encounters)

print("How many regions are there?")
length(unique(annual_region_encounters$Region))


#Pipe the dataframe into the distinct function so that it can pull all the unique regions. 
annual_region_encounters %>%
  distinct(Region)



```
Now we have an idea of what the dataset contains we want to visualize it. 

```{r}
# Clean and filter for 2004–2025 so that we can get a full picture of all the data. 
#Pipe in the data and rename the columns for future readability. 
spaghetti_annual_region_encounters_pre <- annual_region_encounters %>%
  rename(
    Year = `Fiscal\r\nYear`,
    Region = `Region`, 
    Encounters = `Quantity`
  ) %>%
  filter(Year >= 2004 & Year <= 2025)

length(unique(annual_region_encounters$Region))


top_regions <- spaghetti_annual_region_encounters_pre %>%
  group_by(Region) %>%
  summarise(Total = sum(Encounters)) %>%
  arrange(desc(Total)) 

spaghetti_annual_region_encounters_filtered_pre <- spaghetti_annual_region_encounters_pre %>%
  filter(Region %in% top_regions$Region)


print(spaghetti_annual_region_encounters_filtered_pre)
ggplot(spaghetti_annual_region_encounters_pre, aes(x = Year, y = Encounters, color = Region)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Spaghetti Plot: Border Encounters by Region (2004–2025)",
    x = "Year",
    y = "Encounters"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14)
  )
```

## Let's break down the borders? 

- The Air Ports of Entry/ Interior:
  + Location- Nationwide: International Airports. 
  + Consists of Travelers arriving via air. 
  
- The Coastal Border: 
  + Location: U.S. coastlines and maritime entry points.
  + Consists of Travelers arriving from Florida, Puerto Rico, U.S. Virgin Islands, Gulf Coast, and some Atlantic seaports. 
  
- The Northern Land Border: 
  + Location: U.S.–Canada land crossings. 
  + Consists of Travelers arriving from Border stations--Vehicle, rail, and pedestrian crossings-- in states like Washington, Michigan, New York, Vermont, and Maine.
  
- The Southwest Land Border: 
  + Location: U.S.–Mexico land crossings.
  + Consists of Travelers arriving from Texas (El Paso, Laredo, Rio Grande), Arizona (Tucson, Yuma), New Mexico, and Southern California (San Diego).

## With that in mind, what is  this data is telling us? 

```{r}
# Clean and filter for 2004–2025 so that we can get a full picture of all the data. 
#Pipe in the data and rename the columns for future readability. 
spaghetti_annual_region_encounters_pre <- annual_region_encounters %>%
  rename(
    Year = `Fiscal\r\nYear`,
    Region = `Region`, 
    Encounters = `Quantity`
  ) %>%
  filter(Year >= 2014 & Year <= 2025)

length(unique(annual_region_encounters$Region))


top_regions <- spaghetti_annual_region_encounters_pre %>%
  group_by(Region) %>%
  summarise(Total = sum(Encounters)) %>%
  arrange(desc(Total)) %>%
  slice_head(n = 5)

spaghetti_annual_region_encounters_filtered_pre <- spaghetti_annual_region_encounters_pre %>%
  filter(Region %in% top_regions$Region)


print(spaghetti_annual_region_encounters_filtered_pre)
ggplot(spaghetti_annual_region_encounters_pre, aes(x = Year, y = Encounters, color = Region)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Spaghetti Plot: Border Encounters by Region (2014–2025)",
    x = "Year",
    y = "Encounters"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14)
  )
```
## Let's look at the spikes...


The next question to answer is: what is driving these stark fluctuations? 

The graphs reflects the policy changes that have taken place administration to administration. 

The most important changes that took place in the context of immigration policy were: 

-	2002: Creation of DHS
-	2004: Post-9/11 enforcement begins
-	2012: DACA introduced
-	2017: Trump’s executive orders
-	2020: Title 42 and COVID-19 emergency rules
-	2023: Changes under Biden (e.g., expanded TPS)


To visualize the influence of these milestones, we overlaid vertical reference lines on our spaghetti plot using geom_vline() in ggplot2. These vertical lines highlight key years where policy shifts likely influenced the volume of border encounters — helping us better connect data trends with real-world decisions.


```{r}
full_data <- spaghetti_annual_region_encounters_pre

ggplot(full_data, aes(x = Year, y = Encounters, color = Region)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +

  geom_vline(xintercept = 2004, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2012, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2015, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2023, linetype = "dashed", color = "gray30") +

  annotate("text", x = 2004, y = max(full_data$Encounters) * 0.95, 
           label = "DHS Begins", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2012, y = max(full_data$Encounters) * 0.85, 
           label = "DACA", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2015, y = max(full_data$Encounters) * 0.75, 
           label = "Family Surge", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2020, y = max(full_data$Encounters) * 0.65, 
           label = "COVID + Title 42", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2023, y = max(full_data$Encounters) * 0.55, 
           label = "Title 42 Ends", angle = 90, vjust = -0.4, size = 3) +

  labs(
    title = "Spaghetti Plot: Border Encounters by Region (2004–2025)",
    subtitle = "Policy milestones overlaid to reflect spikes and dips in migration trends",
    caption = "Source: KHSM Encounters (OFO)",
    x = "Year",
    y = "Encounters"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14)
  )

```
## Summary

Now that we put all these factors together let's sum up the data. 

Between fiscal years 2004 and 2025, using the KHSM encounter data set published by Department of Homeland Security Key Metric Engine, we pop the question progress of American immigrant lands. Over more than two decades, U.S. attempts to control illegal immigration plus operational deviations from the administration of refugees have modified the volume and spread of border encounters.In the year 2004 was chosen as a starting point for our project because it marks an obvious turning point in U.S. immigration policy and national security policy. 

-As response to the September 11, 2001, terrorist attacks fundamentally affected policy changes when congress passed law establishing Department of Homeland Security (DHS) as an independent entity within the structure that is government itself; thus it absorbed several agencies linked to Immigration's future fate,no less than border areas or customs enforcement. 

-By 2004, collecting procedures had become more unified and reliable, permitting year to year comparisons to be made with confidence, renaming illegal immigration as one of the main hegemonic tasks of national defense, this shift changed the direction and character of immigration administration. Because of issues surrounding public safety,regulatory enforcement with respect to immigration gradually leaned more towards national defense, sayings also tended over growing liberty restrictions on Asian southerners. 

-The data set used in this analysis is KHSM Encounters (OFO), which consists of several worksheets, we examined the Sheet On Monthly Counts that provides data at the monthly level about encounter volumes, sheet Annual By Region that provides complete reports on encounters in each region for a whole year, permitting geographical comparisons over time.Including fiscal year and month, region of encounter, total number of encounters, field office respectively though the last column was not used for analysis.

-We used data setup and cleaning through dplyr to clean and reshaping, the data underwent consistent variable naming and data preference publication, the specific racks included the renaming variables to make them easier to manipulate such as converting Fiscal Year to Year 

-Between 2004 and 2025, rows are filtered pass through when it comes to summary and grouping appointments, we did it by month and region, and both visualizations and analytically tools 

-Additionally, we created clear and effective visualizations of various types using ggplot2. Among other things, the following graphs were produced tho Month ly Averages Plot: A bar and point plot indicates that the average number of appointments is month by month, Year over year, this graph made it possible to identify seasonal patterns in the data which were more encounters happen during dry springs and early summers due not only to natural migrations but increased enforcement efforts.

-We implemented the Spaghetti Plot: from 2004 to 2014 By 2015, the second decade of analysis was under closer observation, and the year 2015 witnessed a sudden upsurge in these encounters, caused by an influx of unaccompanied minors and family units from Central America. In 2020 to 2021 another major trend emerged, that was the COVID-19 pandemic and Title 42 policy had the effect of virtually declaring an end to arrest or apprehension for asylum processing. 

-Entirely apart from this steep decline, the Coastal region was in fact bucking percentages with more frequent or steady encounters, suggesting that other paths of migration existed during lock down years. 

2004: Beginning of DHS enforcement 
2012: DACA known as Deferred Action for Childhood Arrivals 
2015: Central American family surge 
2020: COVID-19 and Title 42 
2023: End of Title 42 under Biden administration , these annotations connect fluctuations in encounter data with real world policy and geopolitical events. They show how immigration numbers are rarely random; they respond to a complex interplay of laws, leadership, and global events.

- Interpretation of Trends, from our analysis, we draw several conclusions: Immigration law enforcement is highly reactive to political priorities and crisis events.The Trump administration 2017–2020 was a period of noticeably strict immigration actions that saw multiple executive orders and restraints on both movements, and COVID-19 brought on a historically unprecedented drop in encounters, followed by a sharp rebound after the pandemic 3 especially after emergency policies were removed. 

-Furthermore, The Coastal region showed distinct behavior that could indicate local patterns of maritime migration or areas where enforcement lacks coverage, these findings confirm that immigration is not simply a matter of numbers. They are representative of who we are as a nation, how we react to fear, what we value in terms of human rights, and how global instability is managed at our borders.

-Our visual and contextual analysis reveals that from 2004 to 2025 one cannot understand patterns in U.S. immigration in isolation, they are directly influenced by changes in administration, global health emergencies, economic crimp times, and humanitarian crises.
