---
title: "Presentation"
author: "Team"
date: "`r Sys.Date()`"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)



library(officer)
library(rmarkdown)

pres <- read_pptx()


#rmarkdown::render("Presentation.Rmd")



pres <- add_slide(pres, layout = "Title Slide", master = "Office Theme")
pres <- ph_with(pres, value = "My Presentation Title", location = ph_location_type(type = "title"))

pres <- add_slide(pres, layout = "Title and Content", master = "Office Theme")
pres <- ph_with(pres, value = "Slide Content", location = ph_location_type(type = "body"))


# Adding a table
df <- data.frame(a = 1:10, b = 11:20)
pres <- add_slide(pres, layout = "Title and Content", master = "Office Theme")
pres <- ph_with(pres, value = df, location = ph_location_type(type = "body"))

# Adding a plot
library(ggplot2)
plot <- ggplot(df, aes(x = a, y = b)) + geom_point()
pres <- add_slide(pres, layout = "Blank", master = "Office Theme")
pres <- ph_with(pres, value = plot, location = ph_location_fullsize())


print(pres, target = "presentation.pptx")

```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

