---
title: "Analysis of US Immigration Policy Impact: 2004–2024"
author: "Group Athens - DATA 2401"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(echo = FALSE)

```

## Introduction

Our project centers around the defining feature of American life by exploring immigration trends in the United States using data collected by U.S. Customs and Border Protection’s (CBP) Office of Field Operations from 2004 to 2024 provided by the U.S. Office of Homeland Security Statistics.  How does U.S. Immigration policy impact CPB Encounters?

Our approach overlays important U.S. Immigration policy events onto an exploratory analysis of CBP Encounter data from 2004 to 2024 to look for trends or inflection points in the data compared to U.S. policy enactments.  The data collected by CBP is provided in annual and monthly formats which we explore. Our analysis uses RStudio for data cleaning, data analysis and visualization on the source data directly without manual operations so the analysis is fully reproducible.

## Data Preparation & Display

The CBP Encounters Key Homeland Security Metric (KHSM) includes any encounter of a removable alien. This includes three sets of data:

- People who unlawfully crossed the border between ports of entry.
- People who entered at a port of entry but are inadmissible.
- People expelled under the Title 42 public health order between March 2020 and May 2023 due to COVID-19

Source: [CBP Encounters - Key Homeland Security Metric](https://ohss.dhs.gov/sites/default/files/2025-01/KHSM%20Encounters%20%28OFO%29%20fy25m11.xlsx)

CBP Encounters data provided by Office of Homeland Security Statistics' begins in 2004 and data collection is on going.  Given the project scope and data continuing to be collected, our first step in preparing the data was to filter the data for analysis to only years for which are fully covered in our dataset so 2025 was excluded from our analysis.  Our analysis starts with three initial steps outlined below:

1. Filter data to years between and including 2004 and 2024
2. Sanitize column headers and rename columns to more sensible nomenclature
3. Summarize and visualize data for exploration

## Exploratory Data Analysis

### Annual Region Encounters
Our analysis starts with CBP Encounters by Annual Region.  We overlay six U.S Immigration policy events corresponding to the year enacted.  We see a steep increase in CBP Encounters in 2021 particularly at *Southwest Land Border* and *Air Ports of Entry/Interior* which seemed like an interesting data trend as it coincided directly to the ending of the U.S. Immigration policy of *Migrant Protection Protocol* also known as **Remain in Mexico**.  Let's explore more.


```{r echo=FALSE, warning=FALSE, fig.width=8, fig.height=4.5}


#-----------------CBP Encounters by US Region: 2004–2024-----------------


#Read Microsoft Excel files
library(readxl)
#Use ggplot
library(ggplot2)

#Use dplyr and silence warnings
library(dplyr, warn.conflicts = FALSE)
options(dplyr.summarise.inform = FALSE)


annual_region_encounters <- read_excel("KHSM Encounters (OFO) fy25m11.xlsx", sheet = "Annual Region")


#Sanitize column headers and rename to more intelligible naming
#Filter to appropriate time frame
annual_region_encounters_sanitize <- annual_region_encounters %>%
  rename(
    Year = 'Fiscal\r\nYear',
    Region = 'Region',
    Encounters = 'Quantity'
  ) %>%
  filter(Year >= 2004 & Year <= 2024)


#Sum encounters by region
annual_region_encounters <- annual_region_encounters_sanitize %>%
  group_by(Region, Year) %>%
  summarize(Total = sum(Encounters))


#Add a factor to help order the legend to match quantity of encounters heirarchy
annual_region_encounters <- annual_region_encounters %>%
  dplyr::filter(Region %in% c('Southwest Land Border', 'Air Ports of Entry/Interior', 'Northern Land Border',  'Coastal Border'))
annual_region_encounters$Region <- factor(annual_region_encounters$Region, levels=c('Southwest Land Border', 'Air Ports of Entry/Interior', 'Northern Land Border', 'Coastal Border'))


#Option to not use scientific notation
options(scipen=999)


#Plot spaghetti plot to contrast regional encounters
ggplot(annual_region_encounters, aes(x = Year, y = Total, color = Region)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +

  #Add US policy events
  geom_vline(xintercept = 2004, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2012, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2015, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2021, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2023, linetype = "dashed", color = "gray30") +

  annotate("text", x = 2004, y = max(annual_region_encounters$Total) * 0.85,
           label = "DHS Begins", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2012, y = max(annual_region_encounters$Total) * 0.85,
           label = "DACA", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2015, y = max(annual_region_encounters$Total) * 0.85,
           label = "Family Surge", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2020, y = max(annual_region_encounters$Total) * 0.85,
           label = "COVID + Title 42", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2021, y = max(annual_region_encounters$Total) * 0.8,
           label = "Remain in Mexico Ends", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2023, y = max(annual_region_encounters$Total) * 0.85,
           label = "Title 42 Ends", angle = 90, vjust = -0.4, size = 3) +

labs(
    title = "CBP Encounters by Region: 2004–2024",
    subtitle = "Policy milestones overlaid to reflect spikes and dips in migration trends",
    caption = "Source: Office of Homeland Security Statistics - Customs & Border Patrol Encounters",
    x = "Year",
    y = "Encounters",
    fill = "US Entry"
  ) +
  theme_minimal() +
  theme(
    panel.border = element_rect(color = 'grey', fill = NA, size = 1),
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(size = 9)
  )

```


### Annual Citizenship (Origin) Encounters
Our next analysis was of the Annual Citizenship dataset which we refer to as Annual Origin.  We could not directly visualize all encounters by all countries of origin so we first summarized and counted all encounters by Origin which we then used to select the top five countries by encounter quantity.  We used this result to directly then filter on a new summary by gathering the top five origins and summarize the result by Origin and Year and found total encounters by Origin and Year.  We find another interesting uptick in encounters around the time when Remain in Mexico ended and of the top five countries of origin, the bottom two which did not spike seem to coincide with the Annual Region Encounters above.  Phillipines and Canada both seem like countries of origin that would not be affected by Remain in Mexico and the data trends as such.  Let's next look at types of encounters by family status.


```{r echo=FALSE, warning=FALSE, fig.width=8, fig.height=4.5}


#-----------------CBP Encounters by Top Origins: 2004–2024-----------------


#Read Microsoft Excel files
library(readxl)
#Use ggplot
library(ggplot2)

#Use dplyr and silence warnings
library(dplyr, warn.conflicts = FALSE)
options(dplyr.summarise.inform = FALSE)


# Load the data
file_path <- "KHSM Encounters (OFO) fy25m11.xlsx"
sheet_name <- "Annual Citizenship"
annual_origin_encounters <- read_excel(file_path, sheet = sheet_name)

#Sanitize column headers and rename to more intelligible naming
#Filter to appropriate time frame
annual_origin_encounters_sanitize <- annual_origin_encounters %>%
  rename(
    Year = 'Fiscal\r\nYear',
    Encounters = 'Quantity',
    Origin = 'Citizenship'
  ) %>%
  filter(Year >= 2004 & Year <= 2024)


# Determine the top countries by total citizenships granted
annual_origin_encounters_filtered <- annual_origin_encounters_sanitize %>%
  group_by(Origin) %>%
  summarise(Total = sum(Encounters)) %>%
  arrange(desc(Total)) %>%
  slice_head(n=5)


# Filter the original data to include only these top countries
top_annual_origin_encounters <- annual_origin_encounters_sanitize %>%
  filter(Origin %in% annual_origin_encounters_filtered$Origin) %>%
  group_by(Origin, Year) %>%
  summarise(Total = sum(Encounters))


#Add a factor to help order the legend to match quantity of encounters heirarchy
top_annual_origin_encounters <- top_annual_origin_encounters %>%
  dplyr::filter(Origin %in% c('Haiti', 'Cuba', 'Mexico',  'Philippines', 'Canada'))
top_annual_origin_encounters$Origin <- factor(top_annual_origin_encounters$Origin, levels=c('Haiti', 'Cuba', 'Mexico',  'Philippines', 'Canada'))


#Plot spaghetti plot to contrast encounters by origin
ggplot(top_annual_origin_encounters, aes(x = Year, y = Total, color = Origin)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  
  #Add US policy events
  geom_vline(xintercept = 2004, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2012, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2015, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2021, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2023, linetype = "dashed", color = "gray30") +

  annotate("text", x = 2004, y = max(top_annual_origin_encounters$Total) * 0.85, 
           label = "DHS Begins", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2012, y = max(top_annual_origin_encounters$Total) * 0.85, 
           label = "DACA", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2015, y = max(top_annual_origin_encounters$Total) * 0.85, 
           label = "Family Surge", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2020, y = max(top_annual_origin_encounters$Total) * 0.85, 
           label = "COVID + Title 42", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2021, y = max(top_annual_origin_encounters$Total) * 0.85, 
           label = "Remain in Mexico Ends", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2023, y = max(top_annual_origin_encounters$Total) * 0.85, 
           label = "Title 42 Ends", angle = 90, vjust = -0.4, size = 3) +
  
  labs(
    title = "CBP Encounters by Top Origins: 2004–2024",
    caption = "Source: Office of Homeland Security Statistics - Customs & Border Patrol Encounters",
    x = "Year",
    y = "Encounters"
  ) +
  theme_minimal() +
  theme(
    panel.border = element_rect(color = 'grey', fill = NA, size = 1),
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14)
  )


```


### Annual Encounters by Family Status

Our analysis turns to Annual Family Status dataset which we summarize by Year and Family status to find total encounters by these categories.  We see again the uptick coinciding with the ending of the Remain in Mexico policy but this time we see also an uptick in the proportion of encounters of Family Unit Individuals.  This uptick does not seem evidently related to ending of Remain in Mexico.

```{r echo=FALSE, warning=FALSE, fig.width=8, fig.height=4.5}


#-----------------CBP Encounters by Top Origins: 2004–2024-----------------



library(readxl)
library(dplyr)
library(ggplot2)

#Use dplyr and silence warnings
library(dplyr, warn.conflicts = FALSE)
options(dplyr.summarise.inform = FALSE)


# Load the data
file_path <- "KHSM Encounters (OFO) fy25m11.xlsx"
sheet_name <- "Annual Family Status"
annual_familystatus_encounters <- read_excel(file_path, sheet = sheet_name)


#Sanitize column headers and rename to more intelligible naming
#Filter to appropriate time frame
annual_familystatus_encounters_sanitize <- annual_familystatus_encounters %>%
  rename(
    Year = 'Fiscal\r\nYear',
    Encounters = 'Quantity',
    Family = 'Family Status'
  ) %>%
  filter(Year >= 2004 & Year <= 2024)


# Summarize the data to get the sum of quantities for each family status by year
annual_familystatus_encounters <- annual_familystatus_encounters_sanitize %>%
  group_by(Year, Family) %>%
  summarize(Total = sum(Encounters))


# Create the bar chart
ggplot(annual_familystatus_encounters, aes(x = Year, y = Total, fill = Family)) +
  geom_bar(stat = 'identity') +
  
  geom_vline(xintercept = 2004, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2012, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2015, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2021, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2023, linetype = "dashed", color = "gray30") +

  annotate("text", x = 2004, y = max(annual_familystatus_encounters$Total) * 0.85, 
           label = "DHS Begins", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2012, y = max(annual_familystatus_encounters$Total) * 0.85, 
           label = "DACA", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2015, y = max(annual_familystatus_encounters$Total) * 0.85, 
           label = "Family Surge", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2020, y = max(annual_familystatus_encounters$Total) * 0.85, 
           label = "COVID + Title 42", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2021, y = max(annual_familystatus_encounters$Total) * 0.85, 
           label = "Remain in Mexico Ends", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2023, y = max(annual_familystatus_encounters$Total) * 0.85, 
           label = "Title 42 Ends", angle = 90, vjust = -0.4, size = 3) +
  
  labs(title = 'CBP Encounters by Family Status: 2004–2024',
       caption = "Source: Office of Homeland Security Statistics - Customs & Border Patrol Encounters",
       x = 'Year',
       y = 'Encounters',
       fill = 'Family Status') +
  theme_minimal() +
  theme(
    panel.border = element_rect(color = 'grey', fill = NA, size = 1),
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14)
  )


```


Diving into the Family Status but instead by Region this time shows that _Family Unit Individuals_ CBP Encounters spike only at entry points of _Air Ports of Entry/Interior_ & _Southwest Land Border_ which seems to support the idea that ending Remain in Mexico seems to have spiked encounters in Regions that would be most impacted by the policy from a geographic standpoint.  We order the facet wrapped plots by highest spike in _Family Unit Individuals_ CBP Encountres to best visualize our findings.


```{r echo=FALSE, warning=FALSE, fig.width=8, fig.height=4.5}


#-----------------Regional (facet) - CBP Encounters by Family Status: 2004–2024----------


#Read Microsoft Excel files
library(readxl)
#Use ggplot
library(ggplot2)

#Use dplyr and silence warnings
library(dplyr, warn.conflicts = FALSE)
options(dplyr.summarise.inform = FALSE)


# Load the data
file_path <- "KHSM Encounters (OFO) fy25m11.xlsx"
sheet_name <- "Annual Rgn EnfAdm Fam Status"
annual_region_famstat_encounters <- read_excel(file_path, sheet = sheet_name)

#Sanitize column headers and rename to more intelligible naming
#Filter to appropriate time frame
annual_region_famstat_encounters_sanitize <- annual_region_famstat_encounters %>%
  rename(
    Year = 'Fiscal\r\nYear',
    Encounters = 'Quantity',
    Family = 'Family Status'
  ) %>%
  filter(Year >= 2004 & Year <= 2024)


#Filter the original data to for analysis by Northern Land Border
annual_region_famstat_encounters_northern <- annual_region_famstat_encounters_sanitize %>%
  filter(Region == c('Northern Land Border'))
#Find Total this Region and then add Region back to dataset
annual_region_famstat_encounters_northern <- annual_region_famstat_encounters_northern %>%
  group_by(Family, Year) %>%
  summarise(Total = sum(Encounters)) %>%
  mutate(Region = 'Northern Land Border')


#Filter the original data to for analysis by Air Ports of Entry/Interior
annual_region_famstat_encounters_airport <- annual_region_famstat_encounters_sanitize %>%
  filter(Region == c('Air Ports of Entry/Interior'))
#Find Total this Region and then add Region back to dataset
annual_region_famstat_encounters_airport <- annual_region_famstat_encounters_airport %>%
  group_by(Family, Year) %>%
  summarise(Total = sum(Encounters)) %>%
  mutate(Region = 'Air Ports of Entry/Interior')


#Filter the original data to for analysis by Southwest Land Border
annual_region_famstat_encounters_southwest <- annual_region_famstat_encounters_sanitize %>%
  filter(Region == c('Southwest Land Border'))
#Find Total this Region and then add Region back to dataset
annual_region_famstat_encounters_southwest <- annual_region_famstat_encounters_southwest %>%
  group_by(Family, Year) %>%
  summarise(Total = sum(Encounters)) %>%
  mutate(Region = 'Southwest Land Border')


#Filter the original data to for analysis by Coastal Border
annual_region_famstat_encounters_coastal <- annual_region_famstat_encounters_sanitize %>%
  filter(Region == c('Coastal Border'))
#Find Total this Region and then add Region back to dataset
annual_region_famstat_encounters_coastal <- annual_region_famstat_encounters_coastal %>%
  group_by(Family, Year) %>%
  summarise(Total = sum(Encounters)) %>%
  mutate(Region = 'Coastal Border')



#Union all Regional analysis to be used in Facet Wrap
annual_region_famstat_encounters_all <- bind_rows(annual_region_famstat_encounters_northern, annual_region_famstat_encounters_airport, annual_region_famstat_encounters_southwest, annual_region_famstat_encounters_coastal)


#Add a factor to help order the facet wrap
annual_region_famstat_encounters_all <- annual_region_famstat_encounters_all %>%
  dplyr::filter(Region %in% c('Southwest Land Border', 'Air Ports of Entry/Interior', 'Northern Land Border',  'Coastal Border'))
annual_region_famstat_encounters_all$Region <- factor(annual_region_famstat_encounters_all$Region, levels=c('Southwest Land Border', 'Air Ports of Entry/Interior', 'Northern Land Border', 'Coastal Border'))


#Add a factor to help order legend
annual_region_famstat_encounters_all <- annual_region_famstat_encounters_all %>%
  dplyr::filter(Family %in% c('Single Adults', 'Family Unit Individuals', 'Unaccompanied Children',  'Accompanied Minors'))
annual_region_famstat_encounters_all$Family <- factor(annual_region_famstat_encounters_all$Family, levels=c('Single Adults', 'Family Unit Individuals', 'Unaccompanied Children', 'Accompanied Minors'))



#Plot spaghetti plot to contrast encounters by origin
ggplot(annual_region_famstat_encounters_all, aes(x = Year, y = Total, color = Family)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  facet_wrap(~ Region) +
  
  #Add US policy events
  geom_vline(xintercept = 2004, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2012, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2015, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2021, linetype = "dashed", color = "gray30") +
  geom_vline(xintercept = 2023, linetype = "dashed", color = "gray30") +

  annotate("text", x = 2004, y = max(annual_region_famstat_encounters_all$Total) * 0.65, 
           label = "DHS Begins", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2012, y = max(annual_region_famstat_encounters_all$Total) * 0.65, 
           label = "DACA", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2015, y = max(annual_region_famstat_encounters_all$Total) * 0.65, 
           label = "Family Surge", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2020, y = max(annual_region_famstat_encounters_all$Total) * 0.65, 
           label = "COVID+Title 42", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2021, y = max(annual_region_famstat_encounters_all$Total) * 0.65, 
           label = "Rem In Mex Ends", angle = 90, vjust = -0.4, size = 3) +
  annotate("text", x = 2023, y = max(annual_region_famstat_encounters_all$Total) * 0.65, 
           label = "Title 42 Ends", angle = 90, vjust = -0.4, size = 3) +


  labs(
    title = "Regional - CBP Encounters by Family Status: 2004–2024",
    caption = "Source: Office of Homeland Security Statistics - Customs & Border Patrol Encounters",
    x = "Year",
    y = "Encounters"
  ) +
  theme_minimal() +
  theme(
    panel.border = element_rect(color = 'grey', fill = NA, size = 1),
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14)
  )


```



We also looked at Monthly CBP Encounter data but did not seem to find evidence that either supports or rejects the idea that ending Remain in Mexico played a seemingly large factor in increased in CBP Encounters.


```{r echo=FALSE, warning=FALSE, fig.height=4, fig.width=8}


#-----------------Avg Monthly Encounters (Total): 2004–2025-----------------

#Read Microsoft Excel files
library(readxl)
#Use ggplot
library(ggplot2)

#Use dplyr and silence warnings
library(dplyr, warn.conflicts = FALSE)
options(dplyr.summarise.inform = FALSE)



# Load the data
file_path <- "KHSM Encounters (OFO) fy25m11.xlsx"
sheet_name <- "Monthly Count"
monthly_encounters <- read_excel(file_path, sheet = sheet_name)


#Sanitize column headers and rename to more intelligible naming
#Filter to appropriate time frame
monthly_encounters_sanitize <- monthly_encounters %>%
  rename(
    Year = 'Fiscal\r\nYear',
    Encounters = 'Quantity'
  ) %>%
  filter(Year >= 2004 & Year <= 2024)
      

#Summarize for Monthly Averages
monthly_encounters <- monthly_encounters_sanitize %>%
  group_by(Month) %>%
  summarize(Month_Mean = mean(Encounters))


#Using a vector to represent the order of my category, we can mutate the categorical field
#to a factor which with the order property allows us to order the x-labels in the plot.
month_order <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
#Abbreviate Month Name
monthly_encounters <- monthly_encounters %>%
  mutate(Month = substr(Month, 4, 6))
#Use factor to order Months for visualization
monthly_encounters <- monthly_encounters %>%
  mutate(Month = factor(Month, levels = month_order, ordered = TRUE))


monthly_encounters %>%
  ggplot(aes(x = Month, y = Month_Mean, fill = Month_Mean)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Avg Monthly Encounters: 2004–2024",
    caption = "Source: Office of Homeland Security Statistics - Customs & Border Patrol Encounters",
    x = "Month",
    y = "Monthly Average",
    fill = "Encounters"
    ) +
  theme_minimal() +
  theme(
    panel.border = element_rect(color = 'grey', fill = NA, size = 1),
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14)
  )



```


## Summary

Our analysis seeks to find data trends in U.S. Customs and Border Protection’s Encounters Key Homeland Security Metric (KHSM) to inform how U.S. Immigration policy impacts this defining feature of American life.  Using annual regional encounter data, annual country of origin encounter data and annaul family status encounter data, we identify a large increase in CBP encounters at ports of entry that are most affected by Remain in Mexico policy ending and we identify that top countries of origin based on encouter quantities that would most likely be encountered at these ports of entry impacted by Remain in Mexico increase as well.  Our conclusion is that the ending of the Remain in Mexico policy caused a significant increase in immigration according to the encounters at specific ports of entry.



