---
title: "TestFileRmd2"
Group Name: Athens 2025
Members: Precious Merenu 
         Claudia Dominguez
         Justin Perez

output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("tidyverse")
library(tidyverse)
#install.packages("readxl")
library(readxl)
```



```{r}
#Save the dataset into a variable. 
child_sum_health_dataset <- 'DQS_NHIS_Children_Summary_Statistics_20250410.csv' 

read_child_sum_health <- read_csv(child_sum_health_dataset)

read_child_sum_health
```

```{r}
#Installing ggplot to visualize the data. 
install.packages("ggplot2", dependencies = T) 

ggplot(data = read_child_sum_health, mapping = aes(x = Topic, y = y_variable, color = Classification)) +
   geom_point(size = 5)

```
#Trying to create a plot of the data. 

p <- ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent)) 
p <- p +  geom_point() + scale_x_log10() 
p


monthly_encounters <- read_excel(encounters, sheet = 'Monthly Count')

monthly_encounter_df <- monthly_encounters %>% mutate(Fiscal_Year='Fiscal Year') %>%
  group_by(Month) %>%
  summarize(Month_Mean = mean(Quantity))

print(monthly_encounter_df)

monthly_encounter_df %>% 
  ggplot(aes(x=Month, y=Month_Mean)) + 
  geom_point() +
  labs(title = "Key Homeland Security Metric - Office of Field Operations",
       x = "Month",
       y = "Monthly Average") +
  theme_minimal()

monthly_encounter_df %>%
  ggplot(aes(x = Month, y=Month_Mean, fill=Month_Mean)) +
  geom_bar(stat = "identity") +
  labs(title = "Key Homeland Security Metric - Office of Field Operations ",
       x = "Month",
       y = "Monthly Average",) +
  theme_minimal()
